Single-cell Type Order Parameters (scTOP): A parameterless, non-stochastic algorithm for tracking cell identity using single-cell atlases

### 2022-10-17 main.tex
first edits

\usepackage{amsmath}
\usepackage{physics}
\usepackage[pdftex]{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\setlength {\marginparwidth }{2cm}
\usepackage{todonotes}
\hypersetup{
    colorlinks,
    citecolor=blue,
    filecolor=black,
    linkcolor=red,
    urlcolor=blue
}

@@

\begin{abstract}
To elucidate the processes of cell-level phenotypic development and reprogramming, scientists are generating an abundance of single-cell RNA-sequencing (scRNA-seq) tissue atlases, lineage tracing studies, and other measurements. Using this high-dimensional, noisy data to identify cell type and understand the dynamics of cell fate transitions remains an ongoing problem. We introduce single-cell Type Order Parameters (scTOP), an epigenetic-landscape-inspired algorithm to track individual cells in the coordinate space of known cell types. Using existing human and mouse datasets that span across the body, we show scTOP reliably identifies cell type and traces the developmental progress of cells. scTOP visualizes the trajectory of lineage tracing experiments by directly comparing each timepoint state to the final fates. By tracking cells in cell type space, scTOP provides the technical foundation to understand in vivo developmental paths and possible bifurcations in the phenotypic landscape.
\end{abstract}

@@

\section{Introduction}
Many adult animals exhibit hundreds of specialized phenotypic cell states, all of which arose from a small number of embryonic cells. In the early stages of development, embryonic cells possess the transcriptomic flexibility to become virtually any of the eventual cell fates. The questions of how and why cells end up in one fate or another is of interest both in biology and the physics of self-organizing systems. Understanding which cell fate transitions are possible is necessary to establish a theoretical framework for development. Understanding transitions is also important for developing directed differentiation protocols to engineer cells in vitro.

An indispensable way to probe cell type is by measuring the transcriptome. Capturing and quantifying the amounts of mRNA in a cell provides a snapshot of its phenotype. With the advent of single-cell RNA-sequencing (scRNA-seq), gene expression data of individual cells are being produced at unprecedented rates. Typical scRNA-seq measurements involve thousands of genes and dozens to millions of cells, resulting in high-dimensional data. Due to the technical challenges of sequencing RNA, the resulting data is noisy and sparse [Lahnemann]. Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero-counts called “dropouts”. Experimental-to-experiment differences in sequencing method, platform, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called “batch effects.” These challenges merit careful consideration of how to normalize and analyze the data in a way that allows the study of biological variation without obfuscation by technical variation.

Three major computational approaches to examining cell fate transitions in scRNA-seq data are clustering, trajectory inference, and automatic cell identification. Clustering involves applying dimensional reduction (such as UMAP or SPRING [citations]) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells. Groups are then manually annotated according to expression levels of marker genes. Trajectory inference analysis arranges cells on a manifold and assigns pseudo-times according to transcriptomic similarity. There are also methods like partition-based graph abstraction [Wolf] that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions. Machine learning classifiers such as support vector machines are able to automatically identify cells with high accuracy [Abdelaal].

Although there are many useful computational tools, the common methods share certain limitations. Dimensional reduction is often one of the first steps to analyze scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships [Chari]. Distorted embeddings inconsistently represent distances between individual data points and between cell types, which makes it difficult to interpret cell fate transitions. This issue carries over to trajectory inference since nearest neighbors can vary substantially according to the embedding. Another limitation is the lack of reproducibility inherent in stochastic machine learning methods [Wattenberg] like t-SNE and UMAP, which produce different results whenever they are executed. Depending on exactly which data points are included, the output distribution may change significantly, even of data points that are retained from one iteration to another. The algorithm results also widely vary depending on input parameters such as perplexity. These parameters insert implicit assumptions about the locality of the data features. Since different papers use different parameters, each use of these algorithms applies different assumptions, rendering quantitative comparison between outputs impossible.

@@

Our algorithm, single-cell Type Order Parameters (scTOP), reproducibly projects scRNA-seq data onto the dimensions of known cell types. Order parameters are variables whose values change as the system’s phase changes. Inspired by the epigenetic landscape and spin glass physics, scTOP treats cell types as phases and calculates the order parameters for transitions between them. It builds on previous work applying attractor-based order parameters to bulk RNA-sequencing [citations]. The inputs into scTOP are the scRNA-seq measurements of query data as well as a reference basis of known cell types. The reference basis contains the archetypal gene expression profiles for the known cell types. scTOP outputs the coordinates of the query data in cell type space; each output dimension corresponds to a measure of similarity between the query data and one of the known cell types in the reference basis. Instead of reducing the data to two arbitrary dimensions, scTOP reduces the data to the number of provided cell types. With scTOP, it’s possible to visualize the path of a cell on axes measuring similarity to the most relevant cell types. The assumptions of the algorithm are explicit: that the known cell types are dynamical system attractors, and that the provided reference basis accurately represents the gene expression profiles of the known cell types. scTOP allows direct comparison between datasets, since the coordinates of and distances between data points do not change from application to application unless the reference basis is changed. scTOP is a robust, deterministic algorithm that creates consistent, intuitive axes on which to observe cell fate transitions.

@@

To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system [cite Huang et al.]. This projection method has been applied previously to bulk RNA-seq samples [citations: Lang et al., 2014; Dame et al., 2017; Pusuluri et al., 2018; Ikonomou et al., 2020a], and we will summarize the theoretical background here. A cell type is a state of the gene regulatory network, so a natural way to represent attractors in this system is with an attractor neural network. If we represent each gene as a node in a network with an associated value measuring gene expression (positive value representing high gene expression, low value indicating low gene expression), then a cell type may be denoted by a vector of gene expression values. Cell types are often highly similar in their patterns of gene regulation. Kanter and Sompolinsky [citation] defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors. For the same system, they also defined order parameters $a^{\mu}$ which are de-correlated versions of the conventional spin glass order parameter, magnetization. $a^{\mu}$ measures the similarity between a given network state and the network state $\mu$, which is a stored pattern (in our case, a gene regulatory pattern that corresponds to a cell type). As [Lang et al.] have shown, the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data. However, the cells digested by bulk RNA-seq may be in different stages of differentiation, and bulk processing renders it impossible to measure cell fate transitions beyond an average tissue state. scTOP uses the same order parameters, $a^{\mu}$, to track cell type at the resolution of individual cells, making it possible to directly observe cells in different stages of differentiation.

In pre-processing, the raw RNA-seq count data are first normalized, then z-scored. An explanation of the process and how it affects the data distributions is provided in Appendix section \ref{preprocessing}.

The way the data is pre-processed before applying the projection method is a vital component of the algorithm, as shown in figure \ref{FIG:1} (a). Each sample is pre-processed independently, unlike other algorithms which process data across samples to find, for example, the most variable genes. Depending on the particular application of scTOP, "sample" may refer to either an individual cell or the average expression levels of a subpopulation of cells. By processing each sample independently, the algorithm output for one sample is independent of which samples are analyzed alongside that one sample. scTOP involves one-shot learning, and there is no stochasticity involved; the algorithm will produce the same output regardless of the particular instance. There are no hyper-parameters involved in the training, which bypasses another source of bias. The only assumptions involved in the algorithm are explicit assumptions: the assumptions that the gene expression profiles included in the reference basis are truly representative of the relevant cell types, and that the cell types included are indeed distinct cell phenotypes.

Constructing a representative, accurate reference basis that is relevant to the data being analyzed is vital to the scTOP algorithm. This process is shown in figure \ref{FIG:1} (b). First, relevant single-cell RNA-seq atlases are gathered to be processed. For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \todo[inline]{Add reference to MCA}, which contains single-cell samples of mouse tissues across \todo[inline]{add \# of organs} organs. Once the relevant atlases are identified, we take an average across each cell population that corresponds to a distinct cell type. It is important that enough cells are included in each average, in order to mitigate the potent effects of noise. The minimum sample size to achieve reasonable results is 200 cells. Noise affects the algorithm most strongly when it appears in the basis, since the de-correlation operation involved in the projection separates cells according to the canonical expression levels included in the reference basis. \todo[inline]{include hill function discussion in appendix to explain how reference sample size affects scTOP score?} After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is pre-processed separately. This creates a modular reference basis where cell types can easily be replaced, removed, or added at will.

@@

\subsection{Benchmarking and validation: Classification of cell fate}
scTOP scores reliably and accurately identify the cell types of individual cells as well as of clusters of cells. To show this, we applied the algorithm to several scRNA-seq datasets across species and across laboratories. scTOP is organism-agnostic as long as an appropriate reference basis is used. Mice and humans are among the most common subjects of scRNA-seq measurements. Many atlases across organs and stages of development have been created for both species. We used some of these atlases, such as the Mouse Cell Atlas [citation] and the Atlas of Lung Development [citation], as reference bases to analyze datasets from various sources.

@@

To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al. [citation] onto the mouse cell atlas. The Herriges et al. b dataset  measures gene expression levels of six different specialized lung types from mice used as control samples. See figure \ref{FIG:2} (a). The reference basis used to find the individual scTOP scores for this data set was derived from the Mouse Cell Atlas. Despite the fact that the Kotton lab dataset and the Mouse Cell Atlas were created in different laboratories and different sequencing methods, the resulting scTOP scores distinctly separate the cells from the Kotton laboratory. The resulting clusters agree with the annotations of the true cell type provided by the Kotton laboratory.

@@

The hematopoietic dataset\cite{weinreb_lineage_2020} represents thousands of hematopoietic stem cells and their cloned sister cells in various states of differentiation. scRNA-seq destroys the cell it measures, but by barcoding cloned cells, this dataset traces the lineage of each clone family. Day 2 undifferentiated cells without clones and day 6 differentiated cells without clones were used to form the reference basis. Then, all of the cloned data was analyzed using scTOP. Figure \ref{FIG:4} (c) shows 2D histograms of every single cloned cell regardless of final fate, colored according to the most likely time for each of the bins (see appendix for details \todo{add info in appendix}). The histograms with the day 2 undifferentiated scTOP scores on the x-axis show a clear path to differentiation, starting at high scores for the undifferentiated type at early days (which is represented by the lighter color) then decreasing undifferentiated-type score as time passes, with some portion ending up with high scores for each of the possible differentiated fates.

@@

\todo{separate undiff plots}

\begin{figure}
	\centering
		\includegraphics[scale=0.5]{figs/fig4.pdf}
	\caption{Beyond bench-marking: tracing development in cell-type space with scTOP. (a) scTOP illustrates specification of alveolar type I and type II cells in murine embryonic development. Alveolar cells from the Atlas of Lung Development are compared to a reference basis constructed from adult lung cells from [Kotton lab citation] and early epithelial cells from the Atlas of Lung Development. The labels next to each data point indicate the age of the mouse embryo from which lungs were extracted, from post-conception day 12 (E12) to day of birth (P0). The points with O-shaped (X-shaped) markers are alveolar type I (II) cells. The x-axis (y-axis) corresponds to the scTOP score for alveolar type I (II) identities. The shade of each point corresponds to the scTOP score for "early epithelium" identity. (b), (c) scTOP shows cell fate specification of hematopoietic cells using lineage tracing data from Weinreb et al. (2020)}
	\label{FIG:4}
\end{figure}

@@

\subsection{Analysis details}
details on how ref. bases were constructed, specific things done for each dataset, etc.

@@

Describe how MCA, TM, and subsets of data were used. number of cells taken.

>>>>

\usepackage{amsmath}
\usepackage{physics}
\usepackage[pdftex]{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\setlength {\marginparwidth }{2cm}
\hypersetup{
    colorlinks,
    citecolor=blue,
    filecolor=black,
    linkcolor=red,
    urlcolor=blue
}

@@

\begin{abstract}
To better understand cell-level phenotypic development and reprogramming, scientists are generating an abundance of single-cell RNA-sequencing (scRNA-seq) tissue atlases, lineage tracing studies, and other measurements. Using this high-dimensional, noisy data to identify cell type and understand the dynamics of cell fate transitions remains an ongoing problem. We introduce single-cell Type Order Parameters (scTOP), an epigenetic-landscape-inspired algorithm to locate individual cells in the coordinate space of known cell types. We use existing human and mouse datasets that span across the body to show scTOP reliably identifies cell type and traces the developmental progress of cells. scTOP visualizes the trajectory of lineage tracing experiments by directly comparing each timepoint state to the final fates. By tracking cells in cell type space, scTOP provides the technical foundation to understand in vivo developmental paths and possible bifurcations in the phenotypic landscape.
\end{abstract}

@@

\section{Introduction}
Many adult animals exhibit hundreds of specialized cell states, all of which arose from a small number of embryonic cells. In early stages of development, embryonic cells possess the transcriptomic flexibility to become virtually any of the eventual cell fates. The questions of how and why cells end up in one fate or another is of interest both in biology and the physics of self-organizing systems. Understanding which cell fate transitions are possible is necessary to establish a theoretical framework for development. Understanding transitions is also important for developing directed differentiation protocols to engineer cells in vitro.

An indispensable way to probe cell type is by measuring the transcriptome. Capturing and quantifying the amounts of mRNA in a cell provides a snapshot of its phenotype. With the advancement of single-cell RNA-sequencing (scRNA-seq), gene expression data of individual cells are being produced at unprecedented rates. Typical scRNA-seq measurements involve thousands of genes and dozens to millions of cells, resulting in high-dimensional data. Due to the technical challenges of sequencing RNA, the resulting data is noisy and sparse \cite{lahnemann_eleven_2020}. Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero-counts called “dropouts”. Experimental-to-experiment differences in sequencing method, platform, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called “batch effects.” These challenges merit careful consideration of how to normalize and analyze the data in a way that permits study of biological variation without obfuscation by technical variation.

clustering, trajectory inference, and automatic cell identification are three major computational approaches to examining cell fate using scRNA-seq data. Clustering involves applying dimensional reduction (such as UMAP \cite{mcinnes_umap_2018} or SPRING \cite{weinreb_spring_2018}) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells. These groups are often manually annotated according to expression levels of marker genes. Trajectory inference analysis arranges cells on a manifold and assigns pseudo-times according to transcriptomic similarity. There are also methods like partition-based graph abstraction \cite{wolf_paga_2019} that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions. Machine learning classifiers such as support vector machines are able to automatically identify cells with high accuracy \cite{abdelaal_comparison_2019}.

Although there are many useful computational tools to extract cell identity from scRNA-seq, the common methods share certain limitations. Dimensional reduction is often one of the first steps to analyze scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships \cite{chari_specious_2021}. Distorted embeddings inconsistently represent distances between individual data points and between cell types, which makes it difficult to interpret cell fate transitions. This issue carries over to trajectory inference, since it uses nearest neighbors which can vary substantially according to the embedding. Another limitation is the lack of reproducibility inherent in stochastic machine learning methods \cite{wattenberg_how_2016} like t-SNE and UMAP, which output different results whenever they are executed. Depending on exactly which data points are included, the output distribution may change significantly, even of data points that are retained from one iteration to another. The algorithm results also widely vary depending on input parameters such as perplexity. These parameters insert implicit assumptions about the locality of the data features. Since different papers use different parameters, each use of these algorithms applies different assumptions, rendering quantitative comparison between outputs impossible.

@@

Our algorithm, single-cell Type Order Parameters (scTOP), reproducibly projects scRNA-seq data onto the dimensions of known cell types. Order parameters are variables whose values change as the system’s phase changes. Inspired by the epigenetic landscape and spin glass physics, scTOP treats cell types as phases and calculates the order parameters for transitions between them. It builds on previous work applying attractor-based order parameters to bulk RNA-sequencing\cite{lang_epigenetic_2014} \cite{dame_thyroid_2017}\cite{pusuluri_cellular_2017} \cite{ikonomou_vivo_2020}. The inputs into scTOP are the scRNA-seq measurements of query data as well as a reference basis of known cell types. The reference basis contains the archetypal gene expression profiles for the known cell types. scTOP outputs the coordinates of the query data in cell type space; each output dimension corresponds to a measure of similarity between the query data and one of the known cell types in the reference basis. Instead of reducing the data to two arbitrary dimensions, scTOP reduces the data to the number of provided cell types. With scTOP, it’s possible to visualize the path of a cell on axes measuring similarity to the most relevant cell types. The assumptions of the algorithm are explicit: that the known cell types are dynamical system attractors, and that the provided reference basis accurately represents the gene expression profiles of the known cell types. scTOP allows direct comparison between datasets, since the coordinates of and distances between data points do not change from application to application unless the reference basis is changed. scTOP is a robust, deterministic algorithm that creates consistent, intuitive axes on which to observe cell fate transitions.

@@

To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system \cite{huang_cell_2005}. This projection method has been applied previously to bulk RNA-seq samples, and we will summarize the theoretical background here \cite{lang_epigenetic_2014}. A cell type is a state of the gene regulatory network, so a natural way to represent attractors in this system is with an attractor neural network. If we represent each gene as a node in a network with an associated value measuring gene expression (positive value representing high gene expression, low value indicating low gene expression), then a cell type may be denoted by a vector of gene expression values. Cell types are often highly similar in their patterns of gene regulation. Kanter and Sompolinsky \cite{kanter_associative_1987} defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors. For the same system, they also defined order parameters $a^{\mu}$ which are de-correlated versions of the conventional spin glass order parameter, magnetization. $a^{\mu}$ measures the similarity between a given network state and the network state $\mu$, which is a stored pattern (in our case, a gene regulatory pattern that corresponds to a cell type). As Lang et al. have shown, the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data. However, the cells digested by bulk RNA-seq may be in different stages of differentiation, and bulk processing renders it impossible to measure cell fate transitions beyond an average tissue state. scTOP uses the same order parameters, $a^{\mu}$, to track cell type at the resolution of individual cells, making it possible to directly observe cells in different stages of differentiation.

In pre-processing, the raw RNA-seq count data are first normalized, then z-scored. The details of the process and its effects on data distributions are described in Appendix section \ref{preprocessing}.

The structure of pre-processing before applying the projection method is a vital component of the algorithm, as shown in figure \ref{FIG:1} (a). Each sample is pre-processed independently, unlike other algorithms which process data across samples to find, for example, the most variable genes. Depending on the particular application of scTOP, "sample" may refer to either an individual cell or the average expression levels of a subpopulation of cells. By processing each sample independently, the output for one sample is independent of which samples are analyzed alongside that one sample. scTOP involves one-shot learning, and there is no stochasticity involved; the algorithm will produce the same output regardless of the particular instance of execution. There are no hyper-parameters involved in the training, which bypasses another source of bias. The only assumptions involved in the algorithm are explicit: that the gene expression profiles included in the reference basis are truly representative of the relevant cell types, and that the cell types included are indeed distinct cell phenotypes.

Constructing a representative, accurate reference basis that is relevant to the data being analyzed is vital to the scTOP algorithm. This process is shown in figure \ref{FIG:1} (b). First, relevant single-cell RNA-seq atlases are gathered to be processed. For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \cite{noauthor_mapping_nodate}, which contains single-cell samples of mouse tissues across all major organs. Once the relevant atlases are identified, we take an average across each cell population that corresponds to a distinct cell type. It is important that enough cells are included in each average, in order to mitigate the potent effects of noise. The minimum sample size to achieve reasonable results is 200 cells. Noise affects the algorithm most strongly when it appears in the basis, since the de-correlation operation involved in the projection separates cells according to the canonical expression levels included in the reference basis. After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is pre-processed separately. This creates a modular reference basis where cell types can easily be replaced, removed, or added at will.

@@

\subsection{Benchmarking and validation: Classification of cell fate}
scTOP scores reliably and accurately identify the cell types of individual cells as well as of clusters of cells. To show this, we applied the algorithm to several scRNA-seq datasets across species and across laboratories. scTOP is organism-agnostic as long as an appropriate reference basis is used. Mice and humans are among the most common subjects of scRNA-seq measurements. Many atlases across organs and stages of development have been created for both species. We used some of these atlases, such as the Mouse Cell Atlas \cite{noauthor_mapping_nodate} and the Atlas of Lung Development \cite{negretti_single-cell_2021}, as reference bases to analyze datasets from various sources.

@@

To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al. \cite{herriges_durable_2022} onto the mouse cell atlas. The Herriges et al. b dataset  measures gene expression levels of six different specialized lung types from mice used as control samples. See figure \ref{FIG:2} (a). The reference basis used to find the individual scTOP scores for this data set was derived from the Mouse Cell Atlas. Despite the fact that the Kotton lab dataset and the Mouse Cell Atlas were created in different laboratories and different sequencing methods, the resulting scTOP scores distinctly separate the cells from the Kotton laboratory. The resulting clusters agree with the annotations of the true cell type provided by the Kotton laboratory.

@@

The hematopoietic dataset\cite{weinreb_lineage_2020} represents thousands of hematopoietic stem cells and their cloned sister cells in various states of differentiation. scRNA-seq destroys the cell it measures, but by barcoding cloned cells, this dataset traces the lineage of each clone family. Day 2 undifferentiated cells without clones and day 6 differentiated cells without clones were used to form the reference basis. Then, all of the cloned data was analyzed using scTOP. Figure \ref{FIG:4} (c) shows 2D histograms of every single cloned cell regardless of final fate, colored according to the most likely time for each of the bins (see appendix\ref{analysis} for details). The histograms with the day 2 undifferentiated scTOP scores on the x-axis show a clear path to differentiation, starting at high scores for the undifferentiated type at early days (which is represented by the lighter color) then decreasing undifferentiated-type score as time passes, with some portion ending up with high scores for each of the possible differentiated fates.

@@

\begin{figure}
	\centering
		\includegraphics[scale=0.5]{figs/fig4.pdf}
	\caption{Beyond bench-marking: tracing development in cell-type space with scTOP. (a) scTOP illustrates specification of alveolar type I and type II cells in murine embryonic development. Alveolar cells from the Atlas of Lung Development are compared to a reference basis constructed from adult lung cells from Herriges et al. and early epithelial cells from the Atlas of Lung Development. The labels next to each data point indicate the age of the mouse embryo from which lungs were extracted, from post-conception day 12 (E12) to day of birth (P0). The points with O-shaped (X-shaped) markers are alveolar type I (II) cells. The x-axis (y-axis) corresponds to the scTOP score for alveolar type I (II) identities. The shade of each point corresponds to the scTOP score for "early epithelium" identity. (b), (c) scTOP shows cell fate specification of hematopoietic cells using lineage tracing data from Weinreb et al.}
	\label{FIG:4}
\end{figure}

@@

\subsection{Analysis details \label{analysis}}
details on how ref. bases were constructed, specific things done for each dataset, etc.

@@

Describe how MCA, TM, and subsets of data were used. number of cells taken. hill function discussion to explain how reference sample size affects scTOP score
